Supplemental Online Materials:\\\textit{Double Standards in Judging Collective Action}

### 2022-06-16 Documents/manuscript.tex
update manuscript

\twocolumn[
  \begin{@twocolumnfalse}
    \maketitle
    \begin{abstract}
      \noindent Collective action is a powerful force driving social
change but, at the same time, often sparks contention about what are
acceptable forms of collective action.
    \end{abstract}
  \end{@twocolumnfalse}
  % \renewcommand{\abstractname}{Statement of Relevance}
  % \begin{@twocolumnfalse}
  %   \maketitle
  %   \begin{abstract}
  %   \noindent 
  %   \end{abstract}
  % \end{@twocolumnfalse}
]

@@

In this paper, we examine double standards in judging collective
action---that is, whether observers judge the same protest actions as
more or less acceptable depending on their own and the protesters' group
memberships (\emph{identity-based double standards}) or on the
protesters' cause and how it aligns with their own ideological positions
(\emph{ideology-based double standards}). In two studies, we first
develop an instrument of 25 controversial protest actions, based in item
response theory, to capture double standards in judging collective
action. In two preregistered experiments, we then use this instrument to
test for double standards in judging protest actions for workers' rights
in the United Kingdom and for or against defunding the police in the
United States. We find evidence for ideology-based, but not for
identity-based, double standards: People judge the same protest actions
to be more acceptable when the protesters' cause aligns with their own
ideological positions---but show no ingroup bias when judging collective
action by ingroup and outgroup members.

@@

Collective action can take many forms. Psychologists distinguish between
\emph{normative} collective action that seeks to achieve a political
goal while conforming to the norms of the existing social system and
\emph{non-normative} collective action that violates those norms (Wright
et al., 1990). Recent research has shown that this distinction matters
for how people respond to collective action. Shuman et al. (2021) showed
that non-normative, non-violent collective action is more effective than
either normative or violent collective action at gaining concessions
from those most resistant to social change. Feinberg et al. (2020)
demonstrated, however, that observers are less supportive of social
movements that use extreme protest actions. Similarly, Teixeira et al.
(2020) found that advantaged-group members perceive non-normative
collective action by disadvantaged-group members as more damaging to
their ingroup's social image and are thus less supportive of such
action. This means that activists face the dilemma that non-normative
collective action might be most effective at gaining concessions but
that it also risks reducing popular support for a cause. Together, these
studies show that the distinction between normative and non-normative
collective action is psychologically and societally consequential.

@@

Just as system justification shapes support for, and opposition to,
collective action, we propose that it results in ideology-based double
standards in what actions people consider acceptable forms of collective
action. First, we hypothesize that people will judge the same protest
actions as more acceptable if the protest supports, rather than
challenges, the system (\emph{Hypothesis 2a}). This hypothesis reflects
the assumption that the motivation to justify the system is, to some
extent, universal. Second, we hypothesize that people will judge the
same protest actions as more acceptable if they endorse
system-justifying beliefs and the protest supports the system or if they
reject system-justifying beliefs and the protest challenges the system
(\emph{Hypothesis 2b}). This hypothesis implies an symmetry in how
people with different ideological orientations judge system-supporting
and system-challenging collective action.

@@

In this research, item response theory serves two purposes. First, item
response theory provides a formal and statistical definition of double
standards in judging collective action as respondents being more
accepting of the same protest actions depending on who the protesters
and what they are protesting. In Experiments 1 and 2, we operationalize
double standards as the differences in \(\theta_{(z)}\) between
conditions which, as \(\theta_{(z)}\) is \(z\)-standardized, correspond
to Cohen's \(d\) effect sizes. Second, item response theory provides a
statistical framework for scale development. In Studies 1 and 2, we
select the most discriminating and informative actions to build an
instrument for capturing double standards in the experiments.

@@

@@

@@

In Experiment 2, we tested five hypotheses. We tested two predictions
derived from social identity theory. First, we hypothesize that
observers are more likely to judge an action as an acceptable means of
protest when the protesters are ingroup members than when they are
outgroup members (Hypothesis 1a). Second, we hypothesize that observers
are more likely to judge an action as an acceptable means of protest
when the goal of the protests would benefit the observers' ingroup
(Hypothesis 1b). We will test two predictions derived from system
justification theory. First, we hypothesize that observers are more
likely to judge an action as an acceptable means of protest when the
goal of the protest is to support rather than challenge the system
(Hypothesis 2a). Second, we hypothesize that observers who endorse
system-justifying beliefs are more likely to consider system-supporting
actions acceptable means of protest-and less likely to consider
system-challenging actions acceptable means of protest-than observers
who do not endorse system-justifying beliefs (Hypothesis 2b). In
addition, we will test an alternative hypothesis. We hypothesize that
observers are more likely to consider actions as acceptable means of
protest when they support the cause of the protesters than when they
oppose their cause (Hypothesis 3). We note that Hypotheses 1a, 1b, and
2a make competing predictions about differences between conditions while
Hypotheses 2b and 3 need not contradict each other or any of the other
hypotheses.

@@

We included additional measures to describe the sample, to describe
reactions to the manipulation, or to use in non-preregistered analyses.
In addition to demographic questions, we asked participants how outraged
they were about recent incidents of police violence against Black
Americans, to what extent they identifed with the protesters described
in the study, and to what extent they identifed with their racial
ingroup (1 = \emph{not at all}, 7 = \emph{very much}). We also asked how
often, if at all, participants had participated in protest actions such
as the ones we had asked about (1 = \emph{never}, 5 = \emph{very often})
and whether they has participated in protests for reforming, defunding
or abolishing the police; against reforming, defunding or abolishing the
police; or in neither. We measured political orientation with a one-item
semantic differential scale: ``People often describe their political
orientation as liberal or conservative. On a scale from liberal to
conservative, where would you position yourself?''; 1 = \emph{liberal},
7 = \emph{conservative}).

@@

We reran Model 3 with political orientation, instead of system
justification, as an alternative predictor variable. Our results
mirrored results from the preregistered analyses: More liberal
participants were more likely to consider protest actions for defunding
the police to be acceptable (\(\beta_{xy} = 0.41, [0.17, 0.62]\)) but
more conservative participants were not more likely to consider protest
actions against defunding the police to be acceptable
(\(\beta_{xy} = -0.16, [-0.41, 0.04]\)). As Figure 5 shows, conservative
participants tended to consider all protest actions to be equally
unacceptable while liberal participants considered the same protest
actions to be more acceptable if protesters rallied around a progressive
cause. Our non-preregistered analyses thus found evidence for
ideology-based double standards in judging collective action with a
different operationalization of political ideology.

@@

\begin{itemize}
\tightlist
\item
  Summary: No evidence for identity-based double standards, but some
  evidence for ideology-based double standards
\item
  Two plausible interpretations: (1) conservatives oppose controversial
  protest (no matter the end) \emph{or} (2) the end justifies the mean
\item
  Re (1) it's plausible but what about recent events? (see public
  polling from grant)
\item
  Re (2) why is there no asymmetry?
\item
  Outline the-end-justifies-the-mean idea, highlighting that just
  supporting a cause is not enough, it needs moralization
\item
  Discuss why high-status group are more accepting of controversial
  protest actions
\item
  Discuss limitations (two cultures, two issues, liberal samples;
  system-defending vs.~system-supporting actions; what about other
  conservative causes?)
\item
  Invite researchers to use our paradigm
\end{itemize}

@@

>>>>

\twocolumn[
  \begin{@twocolumnfalse}
    \maketitle
    \begin{abstract}
      \noindent Collective action is a powerful force driving social
change but often sparks contention about what actions are acceptable
means to effect social change. In four studies (total \emph{N} = 2,246),
we investigated double standards in judging collective action---that is,
whether observers will judge the same protest actions as more acceptable
if the protesters are ingroup members and their cause aligns with the
ingroup's interests (\emph{identity-based double standards}) or if the
protesters' cause aligns with the observers' ideological positions
(\emph{ideology-based double standards}). In two studies, we developed
an instrument of 25 controversial protest actions, based in item
response theory, to measure where people draw the line between
acceptable and unacceptable forms of collective action. In two
preregistered experiments, we used this instrument to test for double
standards in judging collective action for workers' rights in the United
Kingdom and for or against defunding the police in the United States. We
found evidence for ideology-based, but not for identity-based, double
standards: Participants judged the same protest actions to be more
acceptable when the protesters' cause aligned with their own
(system-challenging) ideological position---but participants showed no
ingroup bias when judging collective action by ingroup and outgroup
members. Our findings have theoretical and practical implications for
understanding the often divided response to prominent social movements.
    \end{abstract}
  \end{@twocolumnfalse}
  % \renewcommand{\abstractname}{Statement of Relevance}
  % \begin{@twocolumnfalse}
  %   \maketitle
  %   \begin{abstract}
  %   \noindent 
  %   \end{abstract}
  % \end{@twocolumnfalse}
]

@@

In this paper, we examine double standards in judging collective
action---that is, whether observers judge the same protest actions as
more or less acceptable depending on their own and the protesters' group
memberships (\emph{identity-based double standards}) or on the
protesters' cause and how it aligns with their own ideological positions
(\emph{ideology-based double standards}). In two studies, we first
develop an instrument of 25 controversial protest actions, based in item
response theory, to capture double standards in judging collective
action. In two preregistered experiments, we then use this instrument to
test for double standards in judging protest actions for workers' rights
in the United Kingdom and for or against defunding the police in the
United States. We find evidence for ideology-based, but not for
identity-based, double standards: People judge the same protest actions
to be more acceptable when the protesters' cause aligns with their own
(system-challenging) ideological positions---but do not show ingroup
bias when judging collective action by ingroup and outgroup members.

@@

Collective action can take many forms. Psychologists distinguish between
\emph{normative} collective action that seeks to achieve a political
goal while conforming to the norms of the existing social system and
\emph{non-normative} collective action that violates those norms (Wright
et al., 1990). Recent research has shown that this distinction matters
for how people respond to collective action. Shuman et al. (2021) showed
that non-normative, non-violent collective action is more effective than
either normative or violent collective action at gaining concessions
from those most resistant to social change. Feinberg et al. (2020)
demonstrated, however, that observers are less supportive of social
movements that use extreme protest actions. Similarly, Teixeira et al.
(2020) found that advantaged-group members perceive non-normative
collective action by disadvantaged-group members as more damaging to
their ingroup's social image and are thus less supportive of such
action. Verkuyten et al. (2022) found that, in general, the public tends
to be intolerant of non-normative collective action. This means that
activists face the dilemma that non-normative collective action might be
most effective at gaining concessions but that it also risks reducing
popular support for a cause. Together, these studies show that the
distinction between normative and non-normative collective action is
psychologically and societally consequential.

@@

Just as system justification shapes support for, and opposition to,
collective action, we propose that it results in ideology-based double
standards in what actions people consider acceptable forms of collective
action. First, we hypothesize that people will judge the same protest
actions as more acceptable if the protest supports, rather than
challenges, the system (\emph{Hypothesis 2a}). This hypothesis reflects
the assumption that the motivation to justify the system is, to some
extent, universal. Second, we hypothesize that people will judge the
same protest actions as more acceptable if they endorse
system-justifying beliefs and the protest supports the system or if they
reject system-justifying beliefs and the protest challenges the system
(\emph{Hypothesis 2b}). Notably, this hypothesis implies a symmetry in
how people with different ideological orientations judge
system-supporting and system-challenging collective action.

@@

In this research, item response theory serves two purposes. First, item
response theory provides a statistical framework for scale development.
In Studies 1 and 2, we select the most discriminating and informative
actions to build an instrument for capturing double standards in the
experiments. Second, item response theory provides a formal and
statistical definition of double standards in judging collective action
as respondents being more accepting of the same protest actions
depending on who the protesters and what they are protesting. In
Experiments 1 and 2, we operationalize double standards as the
differences in \(\theta_{(z)}\) between conditions which, as
\(\theta_{(z)}\) is \(z\)-standardized, correspond to Cohen's \(d\)
effect sizes.

@@

In two studies, we developed an instrument of 25 protest actions to
capture double standards in judging collective action.

@@

In Experiment 1, we used the newly developed instrument to test for
double standards in judging collective action for workers' rights in the
United Kingdom. That is, we tested whether people with either
working-class or professional jobs applied different standards when
judging collective action taken by people with either working-class or
middle-class jobs to protest against a fictitious government bill
threatening their rights.

By varying the participants' and the protesters' group memberships, we
tested the preregistered hypothesis that, in line with Hypothesis 1a and
1b, participants would judge the same protest actions as more acceptable
if the protesters were ingroup members protesting for the ingroup's
rights than if they were outgroup members protesting for the outgroup's
rights. By comparing reactions to protesters with working-class or
professional jobs, we tested how observers judge collective action by
lower-status (i.e.~working-class) and higher-status
(professional/middle-class) group members. Assuming that collective
action by lower-status group members is more threatening to the unequal
status than collective action by higher-status group members, we tested
the preregistered hypothesis that, in line with Hypothesis 2a,
participants would judge the same protest actions as more acceptable if
the protesters were from a higher-status group than if they were from a
lower-status group. In non-preregistered analyses, we further explored
whether, in line with Hypothesis 2b, participants would judge the same
protest actions as more or less acceptable depending on their own
ideological positions. In this way, Experiment 1 provided a first test
of the hypothesized identity- and ideology-based double standards in
judging collective action.

@@

@@

We included additional measures to describe the sample, to describe
reactions to the manipulation, or to use in non-preregistered analyses.
In addition to demographic questions, we asked participants how outraged
they were about recent incidents of police violence against Black
Americans, to what extent they identifed with the protesters described
in the study, and to what extent they identifed with their racial
ingroup (1 = \emph{not at all}, 7 = \emph{very much}). We also asked how
often, if at all, participants had participated in protest actions such
as the ones we had asked about (1 = \emph{never}, 5 = \emph{very often})
and whether they has participated in protests for reforming, defunding
or abolishing the police; against reforming, defunding or abolishing the
police; or in neither. We measured political orientation with a one-item
semantic differential scale: ``People often describe their political
orientation as liberal or conservative. On a scale from liberal to
conservative, where would you position yourself?'' (1 = \emph{liberal},
7 = \emph{conservative}).

@@

We reran Model 2 with political orientation, instead of system
justification, as an alternative predictor variable. Our results
mirrored results from the preregistered analyses: More liberal
participants were more likely to consider protest actions for defunding
the police to be acceptable (\(\beta_{xy} = 0.41, [0.17, 0.62]\)) but
more conservative participants were not more likely to consider protest
actions against defunding the police to be acceptable
(\(\beta_{xy} = -0.16, [-0.41, 0.04]\)). As Figure 5 shows, conservative
participants tended to consider all protest actions to be equally
unacceptable while liberal participants considered the same protest
actions to be more acceptable when protesters rallied around a
progressive cause. Our non-preregistered analyses thus replicated the
ideology-based double standards from the preregistered analyses with a
different operationalization of political ideology.

@@

\begin{itemize}
\tightlist
\item
  Summary: No evidence for identity-based double standards, but some
  evidence for ideology-based double standards
\item
  Reject the alternative hypothesis that there is consensus about what
  is an acceptable means of protest.
\item
  Relate our findings to Verkuyten et al. (2022)
\item
  Two plausible interpretations: (1) conservatives oppose controversial
  protest (no matter the end) \emph{or} (2) the end justifies the mean
\item
  Re (1) it's plausible but what about recent events? (see public
  polling from grant)
\item
  Re (2) why is there no asymmetry?
\item
  Outline the-end-justifies-the-mean idea, highlighting that just
  supporting a cause is not enough, it needs moralization
\item
  Discuss why high-status group are more accepting of controversial
  protest actions
\item
  Discuss limitations (two cultures, two issues, liberal samples;
  system-defending vs.~system-supporting actions; what about other
  conservative causes?; ingroup bias might need zero-sum situation to
  succeed)
\item
  Invite researchers to use our paradigm
\end{itemize}

@@

\leavevmode\vadjust pre{\hypertarget{ref-verkuyten_intolerance_2022}{}}%
Verkuyten, M., Adelman, L., \& Yogeeswaran, K. (2022). Intolerance of
transgressive protest actions: The differential roles of deontological
and utilitarian morality. \emph{Personality and Social Psychology
Bulletin}. \url{https://doi.org/10.1177/01461672221099709}

### 2022-06-16 Documents/supplemental-materials.Rmd
update manuscript

Second, we set out to recruit 500 participants from the remaining 687 participants, 250 from the lower-status group and 250 from the higher-status group. After giving their informed consent, participants were randomly assigned to read a vignette about a government bill affecting either people in working-class jobs (*lower-status actors*) or people in professional jobs (*higher-status actors*). Participants in both conditions were instructed to carefully read the vignette and to try to imagine what it would be like if this situation was real. Participants in the lower-status actors condition then read the following introduction:

> The government, though not necessarily the current government, is going to introduce a bill that will mostly affect people in working-class jobs. Working-class jobs, in this case, are jobs done by skilled, semi-skilled, unskilled manual workers or by casual workers. These are jobs that do not usually require a university degree. Other jobs are unlikely to be affected.
Participants in the higher-status actors condition instead read the following introduction:

> The government, though not necessarily the current government, is going to introduce a bill that will mostly affect people in professional jobs. Professional jobs are administrative, managerial, or other jobs that usually require a university degree. Other jobs are unlikely to be affected.
Participants in the two conditions then read this almost identically worded paragraph:

>>>>

Second, we set out to recruit 500 participants from the remaining 687 participants, 250 from the lower-status group and 250 from the higher-status group. After giving their informed consent, participants were randomly assigned to read a vignette about a government bill affecting either people in working-class jobs (*lower-status protesters*) or people in professional jobs (*higher-status protesters*). Participants in both conditions were instructed to carefully read the vignette and to try to imagine what it would be like if this situation was real. Participants in the lower-status protesters condition then read the following introduction:

> The government, though not necessarily the current government, is going to introduce a bill that will mostly affect people in working-class jobs. Working-class jobs, in this case, are jobs done by skilled, semi-skilled, unskilled manual workers or by casual workers. These are jobs that do not usually require a university degree. Other jobs are unlikely to be affected.

\noindent Participants in the higher-status protesters condition instead read the following introduction:

> The government, though not necessarily the current government, is going to introduce a bill that will mostly affect people in professional jobs. Professional jobs are administrative, managerial, or other jobs that usually require a university degree. Other jobs are unlikely to be affected.

\noindent Participants in the two conditions then read this almost identically worded paragraph:

### 2022-06-16 Documents/supplemental-materials.tex
update manuscript

First, we recruited a larger pool of 875 participants. Of these, 475
participants did not have a university degree and placed themselves on
the bottom three ranks of the subjective socio-economic status ladder
(lower-status group) and 400 participants had at least an undergraduate
degree and placed themselves on the top four ranks of the subjective
socio-economic status ladder (higher-status group). Participants
completed a screening survey in which they, among other questions,
answered whether they considered either their current job or the jobs
they had had in the past or would have in the future to be a
working-class job or a middle-class/professional job (1 = working-class
job, 2 = middle-class/professional job, 3 = neither). Before answering
this question, participants read the following explanation:

@@

As preregistered, we excluded participants from the lower-status group
who had responded ``middle-class/professional job'' or ``neither'' to
this question; we excluded participants from the higher-status group who
had responded ``working-class job'' or ``neither'' to this question.
This left 687 participants for the next step of the selection procedure.

Second, we set out to recruit 500 participants from the remaining 687
participants, 250 from the lower-status group and 250 from the
higher-status group. After giving their informed consent, participants
were randomly assigned to read a vignette about a government bill
affecting either people in working-class jobs (lower-status actors) or
people in professional jobs (higher-status actors). Participants in both
conditions were instructed to carefully read the vignette and to try to
imagine what it would be like if this situation was real. Participants
in the lower-status actors condition then read the following
introduction:

\begin{quote}
The government, though not necessarily the current government, is going
to introduce a bill that will mostly affect people in working-class
jobs. Working-class jobs, in this case, are jobs done by skilled,
semi-skilled, unskilled manual workers or by casual workers. These are
jobs that do not usually require a university degree. Other jobs are
unlikely to be affected. Participants in the higher-status actors
condition instead read the following introduction:
\end{quote}

\begin{quote}
The government, though not necessarily the current government, is going
to introduce a bill that will mostly affect people in professional jobs.
Professional jobs are administrative, managerial, or other jobs that
usually require a university degree. Other jobs are unlikely to be
affected. Participants in the two conditions then read this almost
identically worded paragraph:
\end{quote}

>>>>

First, we recruited a larger pool of 875 participants. Of these, 475
participants did not have a university degree and placed themselves on
the bottom three ranks of the subjective socio-economic status ladder
(\emph{lower-status group}) and 400 participants had at least an
undergraduate degree and placed themselves on the top four ranks of the
subjective socio-economic status ladder (\emph{higher-status group}).
Participants completed a screening survey in which they, among other
questions, answered whether they considered either their current job or
the jobs they had had in the past or would have in the future to be a
working-class job or a middle-class/professional job (1 =
\emph{working-class job}, 2 = \emph{middle-class/professional job}, 3 =
\emph{neither}). Before answering this question, participants read the
following explanation:

@@

\noindent As preregistered, we excluded participants from the
lower-status group who had responded ``middle-class/professional job''
or ``neither'' to this question; we excluded participants from the
higher-status group who had responded ``working-class job'' or
``neither'' to this question. This left 687 participants for the next
step of the selection procedure.

Second, we set out to recruit 500 participants from the remaining 687
participants, 250 from the lower-status group and 250 from the
higher-status group. After giving their informed consent, participants
were randomly assigned to read a vignette about a government bill
affecting either people in working-class jobs (\emph{lower-status
protesters}) or people in professional jobs (\emph{higher-status
protesters}). Participants in both conditions were instructed to
carefully read the vignette and to try to imagine what it would be like
if this situation was real. Participants in the lower-status protesters
condition then read the following introduction:

\begin{quote}
The government, though not necessarily the current government, is going
to introduce a bill that will mostly affect people in working-class
jobs. Working-class jobs, in this case, are jobs done by skilled,
semi-skilled, unskilled manual workers or by casual workers. These are
jobs that do not usually require a university degree. Other jobs are
unlikely to be affected.
\end{quote}

\noindent Participants in the higher-status protesters condition instead
read the following introduction:

\begin{quote}
The government, though not necessarily the current government, is going
to introduce a bill that will mostly affect people in professional jobs.
Professional jobs are administrative, managerial, or other jobs that
usually require a university degree. Other jobs are unlikely to be
affected.
\end{quote}

\noindent Participants in the two conditions then read this almost
identically worded paragraph:
